Tool-driven Revolutions in Archaeological Science

### 2019-08-08 analysis/paper/jcaa-paper.Rmd
typos seen by Rev B L 470-517

Although a discipline-wide paradigm shift might be off the table for archaeology, what we can see, though, is that there is a need for methods to share research in a way that is reproducible for others, and that archaeologists are looking for tools to make this possible. Using scripting languages and data repositories to share the scripts is a solution that enjoys increasing popularity in the archaeological research community. It has also expanded the 'trading zone' [@Galison_1997] where archaeologists collaborate with specialists in other fields to develop a shared language to get things done. In using scripting languages archaeologists are working at a trading zone with computer scientists and computational researchers in other fields to coordinate how tools from one domain can be useful in another. We can also see archaeologists in similar trading zones when using GIS, network methods, agent-based models, isotope analysis, remote sensing, and DNA analysis to answer archaeological questions. Archaeologists' embrace of, or turn towards, these methods may also be tool-driven revolutions. However, what makes the use of code unique as a tool-driven revolution is that coding is transcendental, it is an universal method with which we can do all, or which can be central to any, of those previously-mentioned approaches. That is why we believe archaeologists should pay attention to scripted workflows regardless of their usual toolkit and methodological interests.

@@

We acknowledge that learning to use code, if one has not recieved formal instruction, is not an easy task. As @Chrysanthi_2012 illustrate, it requires a certain familiarity with the computational tools to be able to use them not just on a technological basis, but creatively and in full knowledge of their restrictions and ambiguities. The awareness and the appropriate handling of shortcomings and contexts of the data used in an analysis is also a time- and thought-consuming effort. This effort is required at the beginning of each archaeological research, and which may be exacerbated by using multiple data sets from varying sources [cf. @Hugget_2015].

As with data (see Fitzpatrick in this volume), sharing code in public takes a little time to ensure that it is fit for others to read. Our casual observations of the ways in which archaeologists are using and sharing code show a high degree of variability in code style and organisation, indicating that currently most archaeologists are independantly solving problems of how to write and share code through trial and error. This situation of guided variation (learning by trial and error) suggests that using code is not yet widespread enough among archaeologists to propagate due to frequency-dependant biases. It also suggests that the professional benefits (i.e. citation effects) are not yet widely known for code use behaviours to propagate due to content biases, and finally that there are not yet enough highly-visible prestigous researchers using code for this behaviour due to propagate due to model-based biases [@Rendell_et_al_2011; @boyd1988culture; @henrich2007dual]. This suggests to us that there is unrealised potential to improve the efficiency of using and code in archaeology by converging on some widely-agreed upon conventions that save time and investing effort in communicating the professional benefits of using code.

>>>>

Although a discipline-wide paradigm shift might be off the table for archaeology, what we can see, is that there is a need for methods to share research in a way that is reproducible for others, and that archaeologists are looking for tools to make this possible. Using scripting languages and data repositories to share the scripts is a solution that enjoys increasing popularity in the archaeological research community. It has also expanded the 'trading zone' [@Galison_1997] where archaeologists collaborate with specialists in other fields to develop a shared language to get things done. In using scripting languages archaeologists are working at a trading zone with computer scientists and computational researchers in other fields to coordinate how tools from one domain can be useful in another. We can also see archaeologists in similar trading zones when using GIS, network methods, agent-based models, isotope analysis, remote sensing, and DNA analysis to answer archaeological questions. Archaeologists' embrace of, or turn towards, these methods may also be tool-driven revolutions. However, what makes the use of code unique as a tool-driven revolution is that coding is transcendental, it is an universal method with which we can do all, or which can be central to any, of those previously-mentioned approaches. That is why we believe archaeologists should pay attention to scripted workflows regardless of their usual toolkit and methodological interests.

@@

We acknowledge that learning to use code, if one has not received formal instruction, is not an easy task. As @Chrysanthi_2012 illustrate, it requires a certain familiarity with the computational tools to be able to use them not just on a technological basis, but creatively and in full knowledge of their restrictions and ambiguities. The awareness and the appropriate handling of shortcomings and contexts of the data used in an analysis is also a time- and thought-consuming effort. This effort is required at the beginning of each archaeological research, and which may be exacerbated by using multiple data sets from varying sources [cf. @Hugget_2015].

As with data (see Fitzpatrick in this volume), sharing code in public takes a little time to ensure that it is fit for others to read. Our casual observations of the ways in which archaeologists are using and sharing code show a high degree of variability in code style and organisation, indicating that currently most archaeologists are independently solving problems of how to write and share code through trial and error. This situation of guided variation (learning by trial and error) suggests that using code is not yet widespread enough among archaeologists to propagate due to frequency-dependant biases. It also suggests that the professional benefits (i.e. citation effects) are not yet widely known for code use behaviours to propagate due to content biases, and finally that there are not yet enough highly-visible prestigous researchers using code for this behaviour due to propagate due to model-based biases [@Rendell_et_al_2011; @boyd1988culture; @henrich2007dual]. This suggests to us that there is unrealised potential to improve the efficiency of using and code in archaeology by converging on some widely-agreed upon conventions that save time and investing effort in communicating the professional benefits of using code.
